Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels

### 2023-07-11 manuscript_english.tex
Changed small things, italics

That is, we simplify our model (as is done in the physical sciences) to
link through fitness the changes in biological populations and their
structure, observed through population genetics, with patterns of
spatial distribution over time. Thus, the goal is to link biogeographic
patterns and evolutionary processes of populations in a single
conceptual framework.

@@

For a given region \(\mathbf{G}\) in te geography, at any point in time,
we have a set of existing environmental conditions
\(\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}\).
We define \(\vec{e}_i =(e_1, e_2, ..., e_v )_i\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables \(\mathbf{E}(\mathbf{G}, t)\). Then
\(R_0(\vec{e}_i)\) is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the
population's net reproductive rate \(R_0\) is a measure of fitness that
depends on environmental conditions. If we take the set of places in
environmental space where these values are greater than a threshold, we
can define a volume in multidimensional space such that \(R(\vec{e})\)
is always positive and greater than this threshold, that is
\(\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}\). i. e. the set of
environmental conditions for which values of fitness are capable of
sustaining populations. We call
\(\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)\). This is a way of
expressing the fundamental niche as a function of environmental
conditions. So we can correspong each point of geography to
environmental space, with different environmental conditions and
different growth rates. However all sites within the region defined by
\(\mathbf{N}_F\) will have viable populations and positive rates
depending on time because we assume that the environment is dynamic and
each geographic point is static, so for each site in geographic space
this growth rate depends on the environment and therefore is different
at each time: \(R_0 = R_0(\vec{e}(t)) \}\). Thus locally the growth rate
will be related to the local population size as long as it is within the
boundary defined by \(\mathbf{N}_F\).

This equation was proposed by Soberon and Peterson Soberón and Peterson
(2020) to relate demographics to a function of fitness and environment.
In that study they tested the validity with monthly variables and
empirical values on a short time scale. relating life history parameters
and the environmental suitability of niche models.

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_0(\vec{e}_i) > 1\) and \(1\) when
\(R_0\) is evaluated at the expected value of \(\vec {e}(t)\) for all
values of the set \(\mathbf{N}_F\) where \(R_0(\vec{e}_i) > 1\). In a
practical way, instead of \(\mathbf{N}_F\), we approximated
\(\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F\) (the
existing niche) from known presence points in the geography
\(\mathbf{G}\) obtained from emprical observations stored in database,
and map to \(\mathbf{E}\) in order to numerically fitting a convex
volume of the environmental space that enclose the known information.

@@

Once we have an approximation of this volume, we consider that each
point in the geography is static and has a trajectory in the
environmental space and for each trajectory the fitness function changes
and therefore the suitability function changes but not in a homogeneous
way, so each site presents different displacement rates in envrionmental
spaces and as a consequence, different demographic rates. So we can
establish that for each site \(i\) in geographic space \(\mathbf{G}\),
there is a population size \(N_i\) that changes with respect to a
fitness function dependent on time-varying environmental conditions:

@@

where we assume, according to Nei (Nei 1986), that genetic drift within
each subpopulation causes the average heterozygosity between populations
(\(H_s\)) to approach zero (\(F_{st} = 1 - H_s/H_t\)), i.e., for \(t=0\)
\(F_{st} = 0\) and for very large \(t\) the index is \(1\). that is, for
\(t=0\) \(F_{st} = 0\) and for very large \(t\) the index is \(1\). We
have included a constant \(k\) without loss of generalization.

@@

We can infer locally for each population \(N_i\) in the geography expect
to be proportional to the environmental suitability for each time
considered, For simplification in the index we select a \(N_i\)
population in the geography and infer that locally, this population is
proportional to the environmental suitability in each time considered,
i. e. \(N_{j} \propto S_{j}(\vec{e}_{j})\) (each index \(j\) is for the
generation time), as we describe above, so we can asume an harmonic mean
of suitabilities
\(\frac{1}{\tau}\sum_{j = 1}^{\tau} \frac{1}{S_j} = \frac{1}{S}\) and
substitue in the previous equation:

@@

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic \emph{Suitability Prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
\(F_{st}\) is lower are in areas where the prevalence over time is
higher, i.e.~environmental conditions remain suitable for a significant
period of time and therefore no effects of genetic drift of populations
have occurred. Conversely, it is expected that in populations with a
high fixation index value, fitness will not have prevailed over the same
time interval even though conditions currently exist and populations are
present.

@@

\begin{itemize}
\tightlist
\item
  Since it is a medium-sized mammal (0.5 kg) its identification is easy.
\item
  The subspecies are geographically isolated, since they are found in
  pine forests surrounded by desert. Therefore we can consider the
  assumption of isolation by distance (Wright 1943).
\item
  There are sufficient occurrence records for the entire range of the
  known distribution.
\item
  It has low speciation rate (0.25) (Upham, Esselstyn, and Jetz 2019),
  therefore, we assume that its fundamental niche is stable over time.
\item
  There is fixation index information (Bono et al. 2018) for different
  populations of the subspecies, which is useful for using SPA as a way
  to explain genetic differentiation between populations.
\end{itemize}

@@

\hypertarget{fixation-index-as-a-functino-of-spa}{%
\subsubsection{Fixation index as a functino of
SPA}\label{fixation-index-as-a-functino-of-spa}}

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from @Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_{st}\) between this populations in matrix form.
The matrix was transformed into a table as pairs between populations
with index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

@@

\begin{figure}
\centering
\includegraphics{all_figures/figure_4.png}
\caption{Suitability prevalence area. Here we observe the harmonic mean
of suitability for each pixel along the distribucion from 0 to 1. The
bold lines shows the the area when the harmonic mean es above 0.5
threshold (blud line) and 0.9 threshold (black line)}
\end{figure}

@@

\begin{longtable}[]{@{}ccc@{}}
\caption{R squared from linear models (fixation index as a function of
suitability prevalence area). The only R squared with statistical
significance is barberi}\tabularnewline
\toprule\noalign{}
population & r.squared & p.value \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
population & r.squared & p.value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
S. a. aberti Carson-SFW & 0.000 & 0.962 \\
S. a. aberti Coconino-Gila & 0.131 & 0.304 \\
S. a. aberti MT-Zuni & 0.002 & 0.894 \\
S. a. aberti San Juan & 0.000 & 0.952 \\
S. a. barberi & 0.794 & 0.001 \\
S. a. chuscensis E & 0.002 & 0.892 \\
S. a. chuscensis W & 0.018 & 0.713 \\
S. a. ferreus Carson E & 0.040 & 0.581 \\
S. a. ferreus Pike & 0.000 & 0.967 \\
S. a. kaibabensis & 0.005 & 0.841 \\
\end{longtable}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth,height=\textheight]{all_figures/figure_6.png}
\caption{Projection of the \(F_{st}\) model to the geography in the
actual distribution of \emph{S aberti}}
\end{figure}

>>>>

That is, we simplify our model (as is done in the physical sciences) to
link through evolutionary fitness the changes in biological populations and their
structure, observed through population genetics, with patterns of
spatial distribution over time. Thus, the goal is to link biogeographic
patterns and evolutionary processes of populations in a single
conceptual framework.

@@

For a given region \(\mathbf{G}\) in the geography, at any point in time,
we have a set of existing environmental conditions
\(\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}\).
We define \(\vec{e}_i =(e_1, e_2, ..., e_v )_i\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables \(\mathbf{E}(\mathbf{G}, t)\). Then
\(R_0(\vec{e}_i)\) is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the
population's net reproductive rate \(R_0\) is a measure of fitness that
depends on environmental conditions. If we take the set of places in
environmental space where these values are greater than a threshold (\textit{k}), we
can define a volume in multidimensional space such that \(R(\vec{e})\)
is always positive and greater than this threshold, that is
\(\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}\). i. e. the set of
environmental conditions for which values of fitness are capable of
sustaining populations. We call
\(\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)\). This is a way of
expressing the fundamental niche as a function of environmental
conditions. So we can corresponding each point of geography to
environmental space, with different environmental conditions and
different growth rates. However all sites within the region defined by
\(\mathbf{N}_F\) will have viable populations and positive rates
depending on time because we assume that the environment is dynamic and
each geographic point is static, so for each site in geographic space
this growth rate depends on the environment and therefore is different
at each time: \(R_0 = R_0(\vec{e}(t)) \}\). Thus locally the growth rate
will be related to the local population size as long as it is within the
boundary defined by \(\mathbf{N}_F\).

This equation was proposed by Soberón and Peterson
(2020) to relate demographics to a function of fitness and environment.
In that study they tested the validity with monthly variables and
empirical values on a short time scale. relating life history parameters
and the environmental suitability of niche models.

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_0(\vec{e}_i) > 1\) and \(1\) when
\(R_0\) is evaluated at the expected value of \(\vec {e}(t)\) for all
values of the set \(\mathbf{N}_F\) where \(R_0(\vec{e}_i) > 1\). In a
practical way, instead of \(\mathbf{N}_F\), we approximated
\(\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F\) (the
existing niche) from known presence points in the geography
\(\mathbf{G}\) obtained from empirical observations stored in database,
and map to \(\mathbf{E}\) in order to numerically fitting a convex
volume of the environmental space that enclose the known information.

@@

Once we have an approximation of this volume, we consider that each
point in the geography is static and has a trajectory in the
environmental space and for each trajectory the fitness function changes
and therefore the suitability function changes but not in a homogeneous
way, so each site presents different displacement rates in environmental
spaces and as a consequence, different demographic rates. So we can
establish that for each site \(i\) in geographic space \(\mathbf{G}\),
there is a population size \(N_i\) that changes with respect to a
fitness function dependent on time-varying environmental conditions:

@@

where we assume, according to Nei (Nei 1986), that genetic drift within
each subpopulation causes the average heterozygosity between populations
(\(H_s\)) to approach zero (\(F_{st} = 1 - H_s/H_t\)). That is, for
\(t=0\) , \(F_{st} = 0\), and for very large \(t\) the index is \(1\). We
have included a constant \(k\) without loss of generalization.

@@

We can infer locally for each population \(N_i\) in the geography expect
to be proportional to the environmental suitability for each time
considered, For simplification in the index we select a \(N_i\)
population in the geography and infer that locally, this population is
proportional to the environmental suitability in each time considered,
i. e. \(N_{j} \propto S_{j}(\vec{e}_{j})\) (each index \(j\) is for the
generation time), as we describe above, so we can assume an harmonic mean
of suitabilities
\(\frac{1}{\tau}\sum_{j = 1}^{\tau} \frac{1}{S_j} = \frac{1}{S}\) and
substitute in the previous equation:

@@

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic \emph{Suitability Prevalence Area}
\(\mathbf{S} = \{S_i | S_i > s\}\), where \textit{s} is a threshold for the endemic historic area (see below), since we expect that areas where the
\(F_{st}\) is lower are in areas where the prevalence over time is
higher, i.e.~environmental conditions remain suitable for a significant
period of time and therefore no effects of genetic drift of populations
have occurred. Conversely, it is expected that in populations with a
high fixation index value, fitness will not have prevailed over the same
time interval even though conditions currently exist and populations are
present.

@@

\begin{itemize}
\tightlist
\item
  Since it is a medium-sized mammal (0.5 kg) its identification is easy.
\item
  The subspecies are geographically isolated, since they are found in
  pine forests surrounded by desert. Therefore we can consider the
  assumption of isolation by distance (Wright 1943).
\item
  There are sufficient occurrence records for the entire range of the
  known distribution.
\item
  It has low speciation rate (0.25) and \textit{ca.} 5 million year divergence time from other squirrels (Upham, Esselstyn, and Jetz 2019); 
  therefore, we assume that its fundamental niche is stable over time.
\item
  There is fixation index information (Bono et al. 2018) for different
  populations of the subspecies, which is useful for using SPA as a way
  to explain genetic differentiation between populations.
\end{itemize}

@@

\hypertarget{fixation-index-as-a-function-of-spa}{%
\subsubsection{Fixation index as a function of
SPA}\label{fixation-index-as-a-function-of-spa}}

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_{st}\) between this populations in matrix form.
The matrix was transformed into a table as pairs between populations
with index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

@@

\begin{figure}
\centering
\includegraphics{all_figures/figure_4.png}
\caption{Suitability prevalence area of \textit{Sciurus aberti} over the last 120,000 years. Here we observe the harmonic mean
of suitability for each pixel along the distribution from 0 to 1. The
bold lines shows the the area when the harmonic mean es above 0.5
threshold (blue line) and 0.9 threshold (black line)}
\end{figure}

@@

\begin{longtable}[]{@{}ccc@{}}
\caption{R squared from linear models (fixation index as a function of
suitability prevalence area for the localities as listed in Bono et al. 2018). The only R squared with statistical
significance is for \textit{S. a. barberi}.}\tabularnewline
\toprule\noalign{}
population & r.squared & p.value \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
population & r.squared & p.value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textit{S. a. aberti} Carson-SFW & 0.000 & 0.962 \\
\textit{S. a. aberti} Coconino-Gila & 0.131 & 0.304 \\
\textit{S. a. aberti} MT-Zuni & 0.002 & 0.894 \\
\textit{S. a. aberti} San Juan & 0.000 & 0.952 \\
\textit{S. a. barberi} & 0.794 & 0.001 \\
\textit{S. a. chuscensis} E & 0.002 & 0.892 \\
\textit{S. a. chuscensis} W & 0.018 & 0.713 \\
\textit{S. a. ferreus} Carson E & 0.040 & 0.581 \\
\textit{S. a. ferreus} Pike & 0.000 & 0.967 \\
\textit{S. a. kaibabensis} & 0.005 & 0.841 \\
\end{longtable}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth,height=\textheight]{all_figures/figure_6.png}
\caption{Projection of the \(F_{st}\) model to the geography in the
 current distribution of \emph{S. aberti}.}
\end{figure}

Discussion
